---
title: "QC_xenobiotics_statistical_analysis"
author: "Rebecca Matos"
date: "2026-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description:

-   Load RAW QC ***xenobiotics*** dataset

-   Calculate measured concentratio, relative accuracy, intra-day and inter-day precision, and  System Suitability according to WADA rules.

-   Build respective graphs

-   Export processed data

## Setting-up

```{r}
# --- Set Working Directory and Create Output Folders ---
message("Setting up directories...")
if (!dir.exists("output")) {
  dir.create("output", recursive = TRUE)
}

# --- Project Initialization ---
message("Initializing project...")
project_dir <- getwd()
project_name <- "Quality_Control_for_data_quality"

message(paste("Project directory:", project_dir))
message(paste("Project name:", project_name))

```

## Data import

```{r }

message("Loading xenobiotics data file...")

xenob_QC <- read_delim("input/xenob_QC.csv",
                   delim = ";",
                   show_col_types = FALSE)

# Clean names
colnames(xenob_QC) <- trimws(colnames(xenob_QC))

xenob_QC$ratio_mc_is <- as.numeric(gsub(",", ".", xenob_QC$ratio_mc_is))

message(paste("Loaded xenobiotics QC:", nrow(xenob_QC), "features"))

```

## Calculate measured concentration

```{r}

qc_measured_conc <- xenob_QC %>%
  group_by(cmp_number, matrix) %>%
  summarize(
    mean_ratio = mean(ratio_mc_is, na.rm = TRUE),
    sd_ratio   = sd(ratio_mc_is, na.rm = TRUE),
    cv_percent = (sd_ratio / mean_ratio) * 100
  ) %>%
  mutate(across(c(mean_ratio, sd_ratio, cv_percent),
                ~ round(., 2))) %>%
  ungroup()


```

## Relative accuracy across matrices blood and urine (% Bias)

```{r}
# Intensity in BL as true value

bl_reference <- xenob_QC %>%
  filter(matrix == "BL") %>%
  group_by(cmp_number, mc_conc) %>%
  summarise(
    true_ratio = mean(ratio_mc_is, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate in blood and urine

qc_accuracy <- xenob_QC %>%
  filter(matrix != "BL") %>%
  left_join(bl_reference,
            by = c("cmp_number", "mc_conc")) %>%
  group_by(cmp_number, matrix, mc_conc) %>%
  summarise(
    mean_ratio = mean(ratio_mc_is, na.rm = TRUE),
    true_ratio = first(true_ratio),
    accuracy_percent = (mean_ratio / true_ratio) * 100,
    bias_percent = accuracy_percent - 100,
    .groups = "drop"
  ) %>%
  mutate(across(c(mean_ratio, true_ratio, accuracy_percent, bias_percent),
                ~ round(., 2)))

# Result interpretation

qc_accuracy %>%
  mutate(
    pass = case_when(
      abs(bias_percent) <= 15 ~ "PASS",
      TRUE ~ "FAIL"
    )
  ) 
```

Accuracy was assessed by comparing mean analyte/internal standard response ratios in biological matrices (blood and urine) against the reference blank matrix (BL).


### Plots 

```{r}
ggplot(qc_accuracy) +
  geom_bar(aes(x = as.factor(matrix), y = accuracy_percent),
           stat = "identity",
           fill = "dodgerblue4",
           alpha = 0.5) +
  geom_errorbar(aes(x = as.factor(matrix),
                    ymin = accuracy_percent - 5,  # Example: ±5% error, adjust as needed
                    ymax = accuracy_percent + 5),
                width = 0.4,
                colour = "black",
                alpha = 0.9,
                size = 0.5) +
  coord_flip() +
  facet_wrap(vars(cmp_number)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Matrix",
    y = "Accuracy (%)"
  )


qc_measured_conc_filtered <- qc_measured_conc %>%
  filter(matrix %in% c("urine ", "blood"))

ggplot(qc_measured_conc_filtered) +
  geom_bar(aes(x = as.factor(matrix), y = cv_percent),
           stat = "identity",
           fill = "dodgerblue4",
           alpha = 0.5) +
  coord_flip() +
  facet_wrap(vars(cmp_number)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Matrix",
    y = "Coefficient of variance (%)"
  )


```

## Intra-day precision

```{r}

intra_day <- xenob_QC %>%
  group_by(cmp_number, day, matrix) %>%
  summarise(
    mean_ratio = mean(ratio_mc_is),
    sd_ratio = sd(ratio_mc_is),
    cv_percent = (sd_ratio/mean_ratio)*100
  ) %>%
  mutate(across(c(mean_ratio, sd_ratio, cv_percent),
                ~ round(., 2)))


```

### Plot

```{r}
ggplot(intra_day, aes(x = cmp_number, y = cv_percent, fill = matrix)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  theme_minimal() +
  labs(
    x = "Compound",
    y = "Intra-day vari.(%)",
    fill = "Matrix"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Inter-day precision

```{r}

inter_day <- xenob_QC %>%
  group_by(cmp_number, matrix) %>%
  summarise(
    mean_ratio = mean(ratio_mc_is),
    sd_ratio = sd(ratio_mc_is),
    cv_percent = (sd_ratio/mean_ratio)*100
  ) %>%
  mutate(across(c(mean_ratio, sd_ratio, cv_percent),
                ~ round(., 2)))

```

### Plot

```{r}
ggplot(inter_day) +
  geom_bar(aes(x = as.factor(matrix), y = cv_percent),
           stat = "identity",
           fill = "dodgerblue4",
           alpha = 0.5) +
  coord_flip() +
  facet_wrap(vars(cmp_number)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Matrix",
    y = "Coefficient of variance (%)"
  )
```


##  System Suitability

```{r}

system_suitability <- xenob_QC %>%
  group_by(cmp_number, day) %>%
  summarise(
    rt_mean = mean(rt, na.rm = TRUE),
    rt_sd   = sd(rt, na.rm = TRUE),
    rt_cv   = (rt_sd / rt_mean) * 100,
    is_cv   = (sd(is_peak_intensity, na.rm = TRUE) /
               mean(is_peak_intensity, na.rm = TRUE)) * 100,
    .groups = "drop"
  ) %>%
  mutate(across(c(rt_mean, rt_sd, rt_cv, rt_cv, is_cv),
                ~ round(., 2)))
```

##  Output summary

```{r}


summary <-
list(
  qc_accuracy = qc_accuracy,
  Intra_day = intra_day,
  Inter_day = inter_day,
  System_Suitability = system_suitability
)


write_csv(qc_accuracy, "output/qc_accuracy.csv")
write_csv(intra_day, "output/intra_day.csv")
write_csv(inter_day, "output/inter_day.csv")
write_csv(system_suitability, "output/system_suitability.csv")

```


